A re-examination of the role of friction in the original Social Force Model

### 2019-05-23 paper.tex
movi los apendices al main text

The investigation is organized as follows. We first recall the SFM in 
Section~\ref{sfm}, while including the precise definitions for flux, density 
and clustered structures. Section~\ref{simulations} presents our numerical 
simulations for pedestrians moving through corridors. The corresponding results 
are shown in Section~\ref{results}. Our main conclusions are detailed in the 
closing Section~\ref{conclusions}. A complementary simple model for pedestrians 
moving through a corridor has been included in the Appendix. \\

@@

We show that the experimental data (say, the flux-density diagram) can be 
modeled under quasi-stationary conditions in the high density regime. Our 
starting point is the re-examination of the (basic) SFM. We hypothesize that two
control parameters, $\mathcal{A}$ and $\mathcal{K}$ (see \ref{appendix_1}) may handle
the collective motion in dense crowds. We presume that physical contact is
a key feature in dense crowds, despite the fact that other issues may
also contribute to the flow reduction \cite{johansson1}. However, the latter could
be satisfactorily omitted in past research \cite{johansson}, and thus, we will not
attempt to introduce further extensions to the (basic) SFM for the
sake of simplicity.\\

@@

The above observations drive the following conclusion. The maximum attainable velocity $v_{max}$ accomplishes a maximum velocity slope (with its associated friction dissipation). Both the friction with the walls and the friction between the pedestrians appear as relevant magnitudes for properly slowing down the crowd velocity, in order to fit the experimental data. \ref{appendix_2} supports this assertion with a simple example.

We tested the friction coefficient modification in \ref{appendix_2} and we found that the fundamental diagram experiences a qualitatively change when the friction coefficient $\kappa$ is varied. We further performed numerical simulations in the context of the SFM. We call $\kappa_i$ as the friction coefficient of the pedestrian-pedestrian interaction and $\kappa_w$ as the friction coefficient of the pedestrian-wall interaction. Fig.~\ref{fgmodified-w22} shows the flow vs density for different values of $\kappa_i$ and $\kappa_w$.\\

Recall that other authors address the ``congested regime problem" by modifying different aspects of the model. Ref.~\cite{parisi2} imposes zero desired velocity once pedestrians are close enough, Ref.~\cite{johansson} increases the relaxation time in order to slow down the net-time headway, and more recently,  Ref.~\cite{kwak} induce the jamming transition by an attraction. Many of these approaches seem to be equivalent. In \ref{appendix_1} we discuss about how the modification of the relaxation time and the increment of the friction coefficient yield a similar effect, since both affect the same term in the reduced-in-units equation of motion.  The relaxation time, however, also affects the social interactions (see Section \ref{Hypotheses} and \ref{appendix_1}).   \\

In this subsection we have shown that an adequate modification of the friction coefficients yields a fundamental diagram that follows qualitatively the behavior reported through empirical data (say flow reduction for the highest densities). We have also discussed different approaches proposed by other authors in order to overcome this problem. See \ref{appendix_1} for a more detailed discussion.\\

Our simple schematic model (see \ref{appendix_2}) suggests 
that the mismatching problem could be addressed 
alternatively by  modifying the friction coefficient. In light 
of this, we performed numerical simulations increasing the 
value of $\kappa$. We were able to mimic the empirical 
fundamental diagram for sufficiently high values of $\kappa$, 
while keeping the original SFM unchanged (without incorporating neither additional forces nor extra parameters). 
This is actually the major achievement of our investigation. \\

All these phenomena suggest that further research needs to be  
done regarding the friction coefficient. The explored values 
introduced through the investigation, however, should not be considered as 
``empirical'' ones. Its true meaning (within the context of the SFM) is related 
to the other parameters in the model (see \ref{appendix_1}). A real consensus 
on empirical values of $\kappa$ is still missing, to our knowledge. Further analysis are needed to fully explore this issue and are currently under development. \\

\section{\label{appendix_1}The reduced equation of motion}

@@

The role of $\tau$ may be somewhat ambiguous whenever the social 
repulsion becomes negligible with respect to the friction. This may occur if 
some kind of balance exists between neighboring pedestrians in 
symmetrical configurations (\textit{i.e.} in crowded corridors). We may 
hypothesize that the ``key control parameter'' may correspond to either 
$\tau$, or, the friction itself $\kappa$. This is an open question, and a 
first order approach to this matter is outlined in \ref{appendix_2}.\\

\section{\label{appendix_2}A simple model for the corridor}

A toy model for a moving crowd along a corridor is the one represented 
schematically in Fig.~\ref{pasillo}. Pedestrians (circles in 
Fig.~\ref{pasillo}) 
are assumed to be lined up from side to side across the corridor, at any given 
position. Social forces in the $x$-direction are further considered to vanish 
because of translational symmetry. Thus, only the sliding friction is allowed 
to balance the pedestrians own desire. The (reduced) movement equation for 
the $x$-direction according to \ref{appendix_1} and Fig.~\ref{pasillo} 
is

@@

>>>>

The investigation is organized as follows. We first recall the SFM in 
Section~\ref{sfm}, while including the precise definitions for flux, density 
and clustered structures. Section~\ref{simulations} presents our numerical 
simulations for pedestrians moving through corridors. The corresponding results 
are shown in Section~\ref{results}. Our main conclusions are detailed in the 
closing Section~\ref{conclusions}.\\

@@

We show that the experimental data (say, the flux-density diagram) can be 
modeled under quasi-stationary conditions in the high density regime. Our 
starting point is the re-examination of the (basic) SFM. We hypothesize that two
control parameters, $\mathcal{A}$ and $\mathcal{K}$ (see Section \ref{appendix_1}) may handle
the collective motion in dense crowds. We presume that physical contact is
a key feature in dense crowds, despite the fact that other issues may
also contribute to the flow reduction \cite{johansson1}. However, the latter could
be satisfactorily omitted in past research \cite{johansson}, and thus, we will not
attempt to introduce further extensions to the (basic) SFM for the
sake of simplicity.\\

@@

The above observations drive the following conclusion. The maximum attainable velocity $v_{max}$ accomplishes a maximum velocity slope (with its associated friction dissipation). Both the friction with the walls and the friction between the pedestrians appear as relevant magnitudes for properly slowing down the crowd velocity, in order to fit the experimental data. Section \ref{appendix_2} supports this assertion with a simple example.

\subsection{\label{appendix_1}The reduced equation of motion}

@@

The role of $\tau$ may be somewhat ambiguous whenever the social 
repulsion becomes negligible with respect to the friction. This may occur if 
some kind of balance exists between neighboring pedestrians in 
symmetrical configurations (\textit{i.e.} in crowded corridors). We may 
hypothesize that the ``key control parameter'' may correspond to either 
$\tau$, or, the friction itself $\kappa$. This is an open question, and a 
first order approach to this matter is outlined in Section~\ref{appendix_2}.\\

\subsection{\label{appendix_2}A simple model for the corridor}

A toy model for a moving crowd along a corridor is the one represented 
schematically in Fig.~\ref{pasillo}. Pedestrians (circles in 
Fig.~\ref{pasillo}) 
are assumed to be lined up from side to side across the corridor, at any given 
position. Social forces in the $x$-direction are further considered to vanish 
because of translational symmetry. Thus, only the sliding friction is allowed 
to balance the pedestrians own desire. The (reduced) movement equation for 
the $x$-direction according to Section \ref{appendix_1} and Fig.~\ref{pasillo} 
is

@@

We tested the friction coefficient modification in Section \ref{appendix_2} and we found that the fundamental diagram experiences a qualitatively change when the friction coefficient $\kappa$ is varied. We further performed numerical simulations in the context of the SFM. We call $\kappa_i$ as the friction coefficient of the pedestrian-pedestrian interaction and $\kappa_w$ as the friction coefficient of the pedestrian-wall interaction. Fig.~\ref{fgmodified-w22} shows the flow vs density for different values of $\kappa_i$ and $\kappa_w$.\\

Recall that other authors address the ``congested regime problem" by modifying different aspects of the model. Ref.~\cite{parisi2} imposes zero desired velocity once pedestrians are close enough, Ref.~\cite{johansson} increases the relaxation time in order to slow down the net-time headway, and more recently,  Ref.~\cite{kwak} induce the jamming transition by an attraction. Many of these approaches seem to be equivalent. In Section \ref{appendix_1} we discuss about how the modification of the relaxation time and the increment of the friction coefficient yield a similar effect, since both affect the same term in the reduced-in-units equation of motion.  The relaxation time, however, also affects the social interactions (see Sections \ref{Hypotheses} and \ref{appendix_1}).   \\

In this subsection we have shown that an adequate modification of the friction coefficients yields a fundamental diagram that follows qualitatively the behavior reported through empirical data (say flow reduction for the highest densities). We have also discussed different approaches proposed by other authors in order to overcome this problem. See Section \ref{appendix_1} for a more detailed discussion.\\

Our simple schematic model (Section \ref{appendix_2}) suggests 
that the mismatching problem could be addressed 
alternatively by  modifying the friction coefficient. In light 
of this, we performed numerical simulations increasing the 
value of $\kappa$. We were able to mimic the empirical 
fundamental diagram for sufficiently high values of $\kappa$, 
while keeping the original SFM unchanged (without incorporating neither additional forces nor extra parameters). 
This is actually the major achievement of our investigation. \\

All these phenomena suggest that further research needs to be  
done regarding the friction coefficient. The explored values 
introduced through the investigation, however, should not be considered as 
``empirical'' ones. Its true meaning (within the context of the SFM) is related 
to the other parameters in the model (see Section \ref{appendix_1}). A real consensus 
on empirical values of $\kappa$ is still missing, to our knowledge. Further analysis are needed to fully explore this issue and are currently under development. \\
